---
title: "Transparency Appendix from Murphy & Tong Study"
subtitle: "Racial Breakdown of Annual DNA Collection by State: Data Sources and Methodology"
author: "Tina Lasisi & João P. Donadio"
date: today
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| echo: false
#| message: false
#| warning: false

# List of required packages
required_packages <- c(
  "tidyverse",   # Data manipulation and visualization
  "knitr",       # Dynamic report generation
  "plotly",      # Interactive plots
  "DT",          # Interactive tables
  "dplyr",       # Data manipulation
  "stringr",     # String manipulation
  "purrr",       # Functional programming
  "here",        # File path management
  "flextable"    # Table formatting
)

# Function to install missing packages
install_missing <- function(packages) {
  for (pkg in packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      message(paste("Installing missing package:", pkg))
      install.packages(pkg, dependencies = TRUE)
    }
  }
}

# Install any missing packages
install_missing(required_packages)

# Load all packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(DT)
  library(plotly)
  library(dplyr)
  library(stringr)
  library(purrr)
  library(here)
  library(flextable)
})

```

## Executive Summary

This document provides complete transparency regarding the data sources and methodology used to compile racial disparities in DNA collection across U.S. states. The original data was collected in Summer 2017, with most data points from 2013-2016.

### Key Findings

-   **Consistent racial disparities**: Black populations show the highest DNA collection rates relative to their population percentage in nearly all states

-   **Data limitations**: Many states lack comprehensive conviction data, requiring the use of prison admission proxies

-   **Methodological challenges**: Hispanic/Latino populations often uncounted or miscategorized in state data

## Methodology Overview

### Data Collection Period

-   Primary collection: Summer 2017

-   Data years used: Single year per state (2012-2016, varies by availability)

-   Census baseline: 2010 U.S. Census for demographic comparisons

### General Challenges Encountered

1.   **Conviction Data Scarcity**: Most states do not publicly disclose comprehensive felony conviction data

2.   **Prison Admission Proxy**: Prison admissions used as substitute for conviction data in majority of states

3.   **Racial Classification Inconsistencies**:

    -   Many states only report "Black" and "White" categories

    -   Hispanic/Latino often classified as ethnicity rather than race

    -   "Other" category frequently used without specification

4.   **Arrest Data Gaps**: Racial makeup of arrests often unavailable or incomplete

## National Summary Table

```{r}
#| label: tbl-national-summary
#| tbl-cap: "Racial Breakdown of Annual DNA Collection for Each State"
#| echo: true
#| code-fold: true
#| code-summary: "Show summary code"
#| results: hold

# Function to parse DNA collection data from text file
parse_dna_data <- function(file_path) {
  
  # Read the entire file
  text_data <- readLines(file_path, warn = FALSE)
  
  # Remove empty lines
  text_data <- text_data[text_data != ""]
  
  # Initialize list to store parsed data
  parsed_data <- list()
  
  # State abbreviations (for reference)
  state_abbrevs <- c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", 
                     "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", 
                     "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", 
                     "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", 
                     "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY", "DC")
  
  # Function to extract percentage and count from patterns like "46% B (18,253)"
  extract_race_data <- function(text, race_letter) {
    pattern <- paste0("([0-9.]+)%\\s*", race_letter, "\\s*\\(([0-9,]+)\\)")
    matches <- str_match(text, pattern)
    if (!is.na(matches[1])) {
      pct <- as.numeric(matches[2])
      count <- as.numeric(gsub(",", "", matches[3]))
      return(list(pct = pct, count = count))
    }
    return(list(pct = NA, count = NA))
  }
  
  # Function to extract just percentage (for demographics and collection rates)
  extract_percentage <- function(text, race_letter) {
    pattern <- paste0("([0-9.]+)%\\s*", race_letter)
    matches <- str_match(text, pattern)
    if (!is.na(matches[1])) {
      return(as.numeric(matches[2]))
    }
    return(NA)
  }
  
  # Process each line
  i <- 1
  while (i <= length(text_data)) {
    line <- text_data[i]
    
    # Check if line is a state abbreviation
    if (line %in% state_abbrevs) {
      state <- line
      
      # Initialize data structure for this state
      state_data <- list(
        State = state,
        Black_DNA_Pct = NA, Black_DNA_N = NA,
        Hispanic_DNA_Pct = NA, Hispanic_DNA_N = NA,
        Asian_DNA_Pct = NA, Asian_DNA_N = NA,
        Native_American_DNA_Pct = NA, Native_American_DNA_N = NA,
        White_DNA_Pct = NA, White_DNA_N = NA,
        Black_Pop_Pct = NA, Hispanic_Pop_Pct = NA, Asian_Pop_Pct = NA,
        Native_American_Pop_Pct = NA, White_Pop_Pct = NA,
        Black_Collection_Rate = NA, Hispanic_Collection_Rate = NA,
        Asian_Collection_Rate = NA, Native_American_Collection_Rate = NA,
        White_Collection_Rate = NA
      )
      
      # Collect all lines for this state until we hit the next state or end of file
      state_lines <- c()
      i <- i + 1
      while (i <= length(text_data) && !(text_data[i] %in% state_abbrevs)) {
        state_lines <- c(state_lines, text_data[i])
        i <- i + 1
      }
      
      # Combine all lines for this state into one text block
      state_text <- paste(state_lines, collapse = " ")
      
      # Split into sections based on the pattern of % B occurrences
      # We need to identify the three sections: DNA Collection, Demographics, Collection Rates
      
      # Find all "% B" patterns to help identify sections
      b_patterns <- str_locate_all(state_text, "[0-9.]+%\\s*B")[[1]]
      
      if (nrow(b_patterns) >= 1) {
        # First section: DNA Collection (has counts in parentheses)
        dna_section_start <- 1
        dna_section_end <- if(nrow(b_patterns) >= 2) b_patterns[2,1] - 1 else nchar(state_text)
        dna_section <- substr(state_text, dna_section_start, dna_section_end)
        
        # Extract DNA collection data (with counts)
        black_dna <- extract_race_data(dna_section, "B")
        hispanic_dna <- extract_race_data(dna_section, "H")
        asian_dna <- extract_race_data(dna_section, "A")
        native_dna <- extract_race_data(dna_section, "N")
        white_dna <- extract_race_data(dna_section, "W")
        
        state_data$Black_DNA_Pct <- black_dna$pct
        state_data$Black_DNA_N <- black_dna$count
        state_data$Hispanic_DNA_Pct <- hispanic_dna$pct
        state_data$Hispanic_DNA_N <- hispanic_dna$count
        state_data$Asian_DNA_Pct <- asian_dna$pct
        state_data$Asian_DNA_N <- asian_dna$count
        state_data$Native_American_DNA_Pct <- native_dna$pct
        state_data$Native_American_DNA_N <- native_dna$count
        state_data$White_DNA_Pct <- white_dna$pct
        state_data$White_DNA_N <- white_dna$count
      }
      
      if (nrow(b_patterns) >= 2) {
        # Second section: Demographics (percentages only, no parentheses)
        demo_section_start <- b_patterns[2,1]
        demo_section_end <- if(nrow(b_patterns) >= 3) b_patterns[3,1] - 1 else nchar(state_text)
        demo_section <- substr(state_text, demo_section_start, demo_section_end)
        
        # Extract demographic percentages
        state_data$Black_Pop_Pct <- extract_percentage(demo_section, "B")
        state_data$Hispanic_Pop_Pct <- extract_percentage(demo_section, "H")
        state_data$Asian_Pop_Pct <- extract_percentage(demo_section, "A")
        state_data$Native_American_Pop_Pct <- extract_percentage(demo_section, "N")
        state_data$White_Pop_Pct <- extract_percentage(demo_section, "W")
      }
      
      if (nrow(b_patterns) >= 3) {
        # Third section: Collection Rates (percentages only, no parentheses)
        rate_section_start <- b_patterns[3,1]
        rate_section <- substr(state_text, rate_section_start, nchar(state_text))
        
        # Extract collection rate percentages
        state_data$Black_Collection_Rate <- extract_percentage(rate_section, "B")
        state_data$Hispanic_Collection_Rate <- extract_percentage(rate_section, "H")
        state_data$Asian_Collection_Rate <- extract_percentage(rate_section, "A")
        state_data$Native_American_Collection_Rate <- extract_percentage(rate_section, "N")
        state_data$White_Collection_Rate <- extract_percentage(rate_section, "W")
      }
      
      # Add to parsed data
      parsed_data[[length(parsed_data) + 1]] <- state_data
      
      # Don't increment i here since we already did it in the while loop
      i <- i - 1
    }
    
    i <- i + 1
  }
  
  # Convert to data frame
  df <- do.call(rbind, lapply(parsed_data, data.frame))
  
  return(df)
}

data_path <- file.path(here("raw", "foia_pdfs", "MurphyTong_Racial_Breakdown.txt"))

summary_data <- parse_dna_data(data_path)

# Display interactive table
datatable(summary_data, 
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "Complete state-by-state breakdown of DNA collection by race")
```

## Disparity Analysis

```{r}
#| label: fig-disparity-rates
#| fig-cap: "DNA Collection Rates by Race Relative to Population Percentage"
#| echo: true
#| code-fold: true
#| code-summary: "Show disparity analysis code"

# Calculate disparity ratios
disparity_data <- summary_data %>%
  filter(!is.na(Black_Collection_Rate) & !is.na(White_Collection_Rate)) %>%
  mutate(Black_White_Ratio = Black_Collection_Rate / White_Collection_Rate) %>%
  arrange(desc(Black_White_Ratio))

# Create visualization
ggplot(disparity_data %>% head(20), aes(x = reorder(State, Black_White_Ratio), y = Black_White_Ratio)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() +
  labs(title = "Top 20 States: Black-White DNA Collection Disparity Ratio",
       subtitle = "Ratio of collection rates (higher = greater disparity)",
       x = "State",
       y = "Black/White Collection Rate Ratio") +
  theme_minimal() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50")

```

## State-by-State Detailed Methodology

```{r}
#| label: sec-state
#| echo: true
#| code-fold: true
#| code-summary: "Show paragraph extraction code"
#| results: hold

# Pre-define all 50 U.S. states
us_states <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", 
               "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", 
               "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", 
               "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", 
               "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
               "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", 
               "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", 
               "Wisconsin", "Wyoming")

# Read the text file
data_file <- file.path(here("raw", "foia_pdfs", "MurphyTong_States_Paragraphs.txt"))

text_content <- readLines(data_file, warn = FALSE)

# Combine all lines into a single string
full_text <- paste(text_content, collapse = "\n")

# Create a pattern to match state names
state_pattern <- paste0("\\b(", paste(us_states, collapse = "|"), ")\\b")
state_matches <- str_locate_all(full_text, state_pattern)[[1]]

# Extract state sections
state_sections <- list()
state_names <- character()

if (nrow(state_matches) > 0) {
  for (i in 1:nrow(state_matches)) {
    start_pos <- state_matches[i, "start"]
    if (i < nrow(state_matches)) {
      end_pos <- state_matches[i + 1, "start"] - 1
    } else {
      end_pos <- nchar(full_text)
    }
    
    state_name <- substr(full_text, start_pos, state_matches[i, "end"])
    state_content <- substr(full_text, start_pos, end_pos)
    
    state_names <- c(state_names, state_name)
    state_sections <- c(state_sections, state_content)
  }
}

# Function to extract information from each state section
parse_state_section <- function(section, state_name) {
  if (is.na(state_name)) return(NULL)
  
  # Extract legal framework
  legal_framework <- str_extract(section, "Legal Framework:[^\n]+") %>%
    {ifelse(is.na(.), NA, str_remove(., "Legal Framework:") %>% str_trim())}
  
  # Extract collection triggers
  collection_triggers <- str_extract(section, "Collection Triggers:[^\n]+") %>%
    {ifelse(is.na(.), NA, str_remove(., "Collection Triggers:") %>% str_trim())}
  
  # Extract data sources - capture everything until Source URLs or Data Limitations
  data_sources_text <- str_extract(section, "Data Sources:[\\s\\S]*?(?=Source URLs:|Data Limitations:|$)")
  data_source_df <- tibble(data_source_type = NA_character_, data_source_note = NA_character_)
  
  if (!is.na(data_sources_text)) {
    data_sources_text <- str_remove(data_sources_text, "Data Sources:") %>% str_trim()
    # Split by newlines and clean up
    data_source_lines <- str_split(data_sources_text, "\\n")[[1]] %>% 
      str_trim() %>% 
      discard(~ .x == "" | str_detect(.x, "^Source URLs:|^Data Limitations:"))
    
    if (length(data_source_lines) > 0) {
      data_source_df <- map_df(data_source_lines, function(line) {
        if (str_detect(line, ":")) {
          tibble(
            data_source_type = str_extract(line, "^[^:]+") %>% str_trim(),
            data_source_note = str_remove(line, "^[^:]+:") %>% str_trim()
          )
        } else {
          tibble(
            data_source_type = line,
            data_source_note = NA_character_
          )
        }
      })
    }
  }
  
  # Extract source URLs
  source_urls_text <- str_extract(section, "Source URLs:[\\s\\S]*?(?=Data Limitations:|$)")
  source_urls <- character(0)
  
  if (!is.na(source_urls_text)) {
    source_urls_text <- str_remove(source_urls_text, "Source URLs:") %>% str_trim()
    source_url_lines <- str_split(source_urls_text, "\\n")[[1]] %>% 
      str_trim() %>% 
      discard(~ .x == "" | str_detect(.x, "^Data Limitations:"))
    
    if (length(source_url_lines) > 0) {
      source_urls <- source_url_lines
    }
  }
  
  # Extract data limitations - capture everything until next state or end
  data_limitations_text <- str_extract(section, "Data Limitations:[\\s\\S]*?(?=\\b(A|Ala|Alas|Ari|Arka|Cali|Colo|Conn|Del|Flo|Geo|Haw|Ida|Ill|Ind|Iow|Kan|Ken|Lou|Mai|Mar|Mas|Mic|Min|Mis|Mon|Neb|Nev|New|Nor|Ohi|Okl|Ore|Pen|Rho|Sou|Ten|Tex|Uta|Ver|Vir|Was|Wis|Wyo)\\b|$)")
  data_limitations <- NA_character_
  
  if (!is.na(data_limitations_text)) {
    data_limitations_text <- str_remove(data_limitations_text, "Data Limitations:") %>% str_trim()
    data_limitations_lines <- str_split(data_limitations_text, "\\n")[[1]] %>% 
      str_trim() %>% 
      discard(~ .x == "" | str_detect(.x, "^\\b(A|Ala|Alas|Ari|Arka|Cali|Colo|Conn|Del|Flo|Geo|Haw|Ida|Ill|Ind|Iow|Kan|Ken|Lou|Mai|Mar|Mas|Mic|Min|Mis|Mon|Neb|Nev|New|Nor|Ohi|Okl|Ore|Pen|Rho|Sou|Ten|Tex|Uta|Ver|Vir|Was|Wis|Wyo)\\b"))
    
    if (length(data_limitations_lines) > 0) {
      data_limitations <- paste(data_limitations_lines, collapse = "; ")
    }
  }
  
  # If no data sources were found, create at least one row for the state
  if (nrow(data_source_df) == 0) {
    data_source_df <- tibble(data_source_type = NA_character_, data_source_note = NA_character_)
  }
  
  # Create result dataframe
  result_df <- data_source_df %>%
    mutate(
      state = state_name,
      legal_framework = legal_framework,
      collection_triggers = collection_triggers,
      source_url = ifelse(length(source_urls) > 0, paste(source_urls, collapse = "; "), NA_character_),
      data_limitations = data_limitations,
      .before = everything()
    )
  
  return(result_df)
}

# Parse all state sections
state_data <- map2_df(state_sections, state_names, parse_state_section, .progress = TRUE)

# Clean up the data - remove rows where all data columns are NA
final_df <- state_data %>%
  mutate(across(where(is.character), ~ ifelse(.x == "" | is.na(.x), NA, .x))) %>%
  filter(!(is.na(data_source_type) & is.na(source_url) & is.na(data_limitations)))

# Fill in the missing legal_framework and other metadata for each state
final_df_clean <- final_df %>%
  group_by(state) %>%
  fill(legal_framework, collection_triggers, .direction = "downup") %>%
  ungroup() %>%
  # Remove rows where legal_framework is NA (these seem to be incomplete entries)
  filter(!is.na(legal_framework))

# Now let's fill source_url and data_limitations across all rows for each state
final_df_clean <- final_df_clean %>%
  group_by(state) %>%
  mutate(
    source_url = ifelse(all(is.na(source_url)), NA, 
                       paste(na.omit(unique(source_url)), collapse = "; ")),
    data_limitations = ifelse(all(is.na(data_limitations)), NA,
                             paste(na.omit(unique(data_limitations)), collapse = "; "))
  ) %>%
  ungroup()

# Create categorization columns for easier analysis
final_df_clean <- final_df_clean %>%
  mutate(
    # Categorize collection triggers
    collection_trigger_category = case_when(
      str_detect(collection_triggers, "(?i)all felony.*convictions.*arrests.*all felonies") ~ "Comprehensive: All felonies + broad arrests",
      str_detect(collection_triggers, "(?i)all felony.*convictions.*arrests.*certain|specific") ~ "Selective: All felonies + specific arrests",
      str_detect(collection_triggers, "(?i)all felony.*convictions.*arrests") ~ "Broad: All felonies + various arrests",
      str_detect(collection_triggers, "(?i)all felony.*convictions") ~ "Felony convictions only",
      str_detect(collection_triggers, "(?i)felony.*misdemeanor.*convictions") ~ "Mixed: Felony + misdemeanor convictions",
      TRUE ~ "Other/Unspecified"
    ),
    
    # Categorize data limitations
    data_limitation_category = case_when(
      is.na(data_limitations) ~ "No limitations noted",
      str_detect(data_limitations, "(?i)no direct.*conviction data") ~ "Missing conviction data",
      str_detect(data_limitations, "(?i)prison admissions.*proxy") ~ "Prison data as proxy",
      str_detect(data_limitations, "(?i)hispanic|ethnicity") ~ "Ethnicity categorization issues",
      str_detect(data_limitations, "(?i)racial.*limited|black.*white.*only") ~ "Limited racial categories",
      str_detect(data_limitations, "(?i)no.*data.*available|unavailable") ~ "Various data unavailable",
      TRUE ~ "Other limitations"
    ),
    
    # Categorize data source types
    data_source_category = case_when(
      str_detect(data_source_type, "(?i)conviction") ~ "Conviction Data",
      str_detect(data_source_type, "(?i)arrest") ~ "Arrest Data",
      str_detect(data_source_type, "(?i)sex|sexual") ~ "Sex Crime Data",
      str_detect(data_source_type, "(?i)prison|admission|correction") ~ "Prison/Incarceration Data",
      TRUE ~ "Other Data Source"
    )
  )

# Create a summary table for quick overview
summary_table <- final_df_clean %>%
  distinct(state, legal_framework, collection_triggers, collection_trigger_category, 
           data_limitations, data_limitation_category, source_url) %>%
  arrange(state)

# Create a categorized view of data sources
data_source_summary <- final_df_clean %>%
  filter(!is.na(data_source_type)) %>%
  group_by(state, data_source_category) %>%
  summarize(
    data_sources = paste(unique(data_source_type), collapse = "; "),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = data_source_category,
    values_from = data_sources,
    values_fill = NA
  )

# Join the summary information
final_summary <- summary_table %>%
  left_join(data_source_summary, by = "state") %>%
  select(state, legal_framework, collection_trigger_category, data_limitation_category)

# Create interactive tables for exploration
datatable_summary <- datatable(
  final_summary,
  extensions = c('Buttons', 'ColReorder', 'Scroller'),
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'colvis'),
    scrollX = TRUE,
    scrollY = "600px",
    scroller = TRUE,
    pageLength = 10,
    columnDefs = list(
      list(className = 'dt-left', targets = 0:(ncol(final_summary)-1)),
      # Only apply width styling to columns that actually exist
      list(width = '200px', targets = c(1, 2, 3)) # Adjust based on your actual column count
    ),
    autoWidth = TRUE
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe hover'
) %>%
  formatStyle(
    columns = names(final_summary),
    fontSize = '12px',
    lineHeight = '90%'
  )

# Show the tables
datatable_summary

```
